Reconstructing the Lensing Mass in the Universe \\ from Photometric Catalogue Data

### 2013-03-01 doc/paper-lcr.tex
modified rest

High precision cosmological distance measurements towards individual objects
such as time delay gravitational lenses or type Ia supernovae are affected by
weak lensing perturbations by galaxies and groups along the line of sight. In
time delay gravitational lenses, ``external convergence,'' $\kappax$, can
dominate the uncertainty in the inferred distances and hence cosmological parameters.
 In this paper we attempt
to reconstruct $\kappax$, due to line of sight structure, using a simple halo
model.
%
We use mock catalogues from the Millennium Simulation, and calibrate and 
compare our reconstructed $\pr(\kappax)$ to ray-traced $\kappax$ ``truth''
values; taking into account realistic uncertainties on redshift and stellar
masses. \tom{We find that the reconstruction of $\kappax$ provides an improvement in
precision of $\sim$50\% over galaxy number counts.} 
%
We find that the lowest-$\kappax$ lines of sight have the best constrained 
$\pr(\kappax)$. In anticipation of future samples with thousands of lenses,
we find that selecting the third of the systems with the
highest precision $\kappax$ estimates gives a  sub-sample of unbiased time
delay distance measurements \phil{with (on average) just} 1\% uncertainty due
to line of sight external convergence effects. Photometric data alone are
sufficient to pre-select the best-constrained lines of sight, and can be done
before investment in lightcurve monitoring. 
%
\phil{Conversely, we show that selecting lines of sight
with high external shear could, with the reconstruction model presented here,
induce biases of up to 1\% in time delay distance.}  We suggest areas for the
improvement of this general analysis framework (including more sophisticated
treatment of high mass structures) that should allow yet more accurate
cosmological inferences to be made.

@@

This paper is organized as follows. We review the relevant gravitational
lensing theory in~\Sref{sec:theory}, and the Millennium Simulation
ray-traced convergences and mock catalogues in \Sref{sec:MS}, before
introducing our simple reconstruction  model in \Sref{sec:model}. We
then test this model in two phases: first, in \Sref{sec:knownMh+z}, with
known redshift and halo mass for every galaxy in a lens field, in order
to quantify the irreducible uncertainty due to unseen mass, and second,
in \Sref{sec:obsMstar+z}, with realistic observational uncertainties on
the observed galaxies' stellar masses and redshifts. In
\Sref{sec:biases} we investigate the potential systematic error induced
by selecting a subset of lines of sight. We discuss our results in
\Sref{sec:discuss} before concluding in \Sref{sec:conclude}.

@@

\section{Systematic Errors due to using an incorrect Stellar Mass--Halo Mass Relation}
\label{sec:biases}

@@

We test the potential size of this systematic error by using the \citet{BehrooziEtal2010} relation to generate the stellar masses of our calibration lines of sight, but the stellar mass--halo mass relation of \citet{MosterEtal2010} for the \"real\" lines of sight that are reconstructed.

@@

Despite the systematic errors the photometric reconstruction is still sufficient to choose the best lines of sight for spectroscopic follow-up. With targetted spectroscopy the systematic error decreases to $\sim 0.014$ for the ensemble, $\sim 0.005$ for the low kappa sample and $\sim 0.018$ for the high shear sample. This result provides further motivation for prioritizing the lenses that reside on the underdense lines of sight.
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\section{Improving precision by including additional information}
%\notes{placeholder for shear}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In this work, we have used  the \MS as a calibration tool. If the real
Universe is not like the \MS, then this calibration will be incorrect.
Repeating this study using a different simulation to make the test catalogues
would allow us to quantify the sensitivity of our results to the simulation
used. Similarly the use of a different semi-analytic model to paint galaxies
onto halos, and an alternative stellar mass--halo mass relation would both
enable similar exploration of these systematic errors. \citet{SuyuEtal2010}
used the relative over-density to mitigate this effect, but it does not
completely remove the dependence on simulation. \new{Throughout this work, we assumed
that the stellar mass--halo mass relation used for the reconstruction is the stellar mass--halo mass
relation followed by the real universe. We discarded the stellar masses from the Millenium Catalogue
due to a large number of excessively stripped satellite galaxies, however if we use these stellar masses
there is no systematic change in the ensemble, but reconstructed $\kappax$ differs by $\sim 0.015$ for individual lines of sight. The effect is smaller ($\sim 0.008$)
for the lowest $\kappax$ lines of sight, and larger ($\sim 0.02$) for high shear lines of sight. This 
change of stellar masses is extreme but demonstrates that using the wrong stellar mass--halo mass relation
can significantly degrade the quality of inference.} Future work should include
investigating the robustness of the reconstruction to these systematic errors.

@@

>>>>

High precision cosmological distance measurements towards individual objects
such as time delay gravitational lenses or type Ia supernovae are affected by
weak lensing perturbations by galaxies and groups along the line of sight. In
time delay gravitational lenses, ``external convergence,'' $\kappax$, can
dominate the uncertainty in the inferred distances and hence cosmological parameters.
 In this paper we attempt
to reconstruct $\kappax$, due to line of sight structure, using a simple halo
model.
%
We use mock catalogues from the Millennium Simulation, and calibrate and 
compare our reconstructed $\pr(\kappax)$ to ray-traced $\kappax$ ``truth''
values; taking into account realistic uncertainties on redshift and stellar
masses. \tom{We find that the reconstruction of $\kappax$ provides an improvement in
precision of $\sim$50\% over galaxy number counts.} 
%
We find that the lowest-$\kappax$ lines of sight have the best constrained 
$\pr(\kappax)$. In anticipation of future samples with thousands of lenses,
we find that selecting the third of the systems with the
highest precision $\kappax$ estimates gives a  sub-sample of unbiased time
delay distance measurements \phil{with (on average) just} 1\% uncertainty due
to line of sight external convergence effects. Photometric data alone are
sufficient to pre-select the best-constrained lines of sight, and can be done
before investment in lightcurve monitoring. 
%
\phil{Conversely, we show that selecting lines of sight
with high external shear could, with the reconstruction model presented here,
induce biases of up to 1\% in time delay distance.} \new{We find that a major potential source of systematic error is uncertainty in the high mass end of the stellar mass-halo mass relation; this could introduce ~2\% biases on the time-delay distance if completely ignored.} We suggest areas for the
improvement of this general analysis framework (including more sophisticated
treatment of high mass structures) that should allow yet more accurate
cosmological inferences to be made.

@@

This paper is organized as follows. We review the relevant gravitational
lensing theory in~\Sref{sec:theory}, and the Millennium Simulation
ray-traced convergences and mock catalogues in \Sref{sec:MS}, before
introducing our simple reconstruction  model in \Sref{sec:model}. We
then test this model in two phases: first, in \Sref{sec:knownMh+z}, with
known redshift and halo mass for every galaxy in a lens field, in order
to quantify the irreducible uncertainty due to unseen mass, and second,
in \Sref{sec:obsMstar+z}, with realistic observational uncertainties on
the observed galaxies' stellar masses and redshifts. In
\Sref{sec:biases} we investigate the potential systematic error induced
by selecting a subset of lines of sight. In
\Sref{sec:SHAMfail} we investigate systematic errors caused by assuming an incorrect stellar-mass relation. We discuss our results in
\Sref{sec:discuss} before concluding in \Sref{sec:conclude}.

@@

\section{Systematic Errors due to using an incorrect Stellar Mass--Halo Mass Relation}
\label{sec:SHAMfail}

@@

We test the potential size of this systematic error by using the \citet{BehrooziEtal2010} relation to generate the stellar masses of our calibration lines of sight, but the stellar mass--halo mass relation of \citet{MosterEtal2010} for the ``real'' lines of sight that are reconstructed.

@@

Despite the systematic errors the photometric reconstruction is still sufficient to choose the best lines of sight for spectroscopic follow-up. With targetted spectroscopy the systematic error decreases to $\sim 0.014$ for the ensemble, $\sim 0.005$ for the low kappa sample and $\sim 0.018$ for the high shear sample. This result provides further motivation for prioritizing the lenses that reside on the underdense lines of sight.

The systematic errors reported herein may be overly pessimistic; the bias is due to differences in the stellar masses predicted for high mass halos. If observations can discriminate between the high mass end of the \citet{BehrooziEtal2010} and \citet{MosterEtal2010} relations then the systematic error on the reconstruction will decrease.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Improving precision by including additional information}
%\notes{placeholder for shear}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In this work, we have used  the \MS as a calibration tool. If the real
Universe is not like the \MS, then this calibration will be incorrect.
Repeating this study using a different simulation to make the test catalogues
would allow us to quantify the sensitivity of our results to the simulation
used. Similarly the use of a different semi-analytic model to paint galaxies
onto halos would enable similar exploration of these systematic errors. 
\citet{SuyuEtal2010}
used the relative over-density to mitigate the effect of simulatioan and
simulation, but it does not completely remove the dependence on simulation.
\new{We have tested the effect of using a different stellar mass--halo mass relation
and found that this can induce significant systematic biases if the true relation is unknown.} Future work should include
investigating the robustness of the reconstruction to these systematic errors \new{and develope proscriptions that mitigate against them}

@@

\item \new{Reconstructing a line of sight with an incorrect stellar mass--halo mass relation can introduce systematic errors on the inferrred time-delay distance at the percent level. This error can only be reduced by tightening the observational limits on the stellar mass--halo mass relation at the high mass end.}
